A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-15 paper/paper.tex
Add auxiliary functions for substitution

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Yiyun Liu}
\orcid{0009-0006-8717-2498}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \country{USA}
}
\email{liuyiyun@seas.upenn.edu}

@@

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

%         \mathit{Renaming} \\
%         [[i {xi}]] & := & [[xi i]] \\
%         [[(Set i) {xi}]] & := & [[Set i]] \\
%         [[Void {xi}]] & := & [[Void]]\\
%         [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
%         [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
%         [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
%         [[Bool {xi}]] & := & [[Bool]] \\
%         [[true {xi}]] & := & [[true]] \\
%         [[false {xi}]] & := & [[false]] \\
%         [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
%         [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
%         [[refl {xi}]] & := & [[refl]] \\
%         [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\

%         \mathit{Shift} \\
%         [[up n A]] & := & A \langle add(n) \rangle \\ \\

%         \mathit{Shift\ one} \\
%         [[up1 A]] & := & \uparrow^1 [[A]] \\ \\

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We will use
the unscoped de Bruijn representation for the rest of the paper,
adopting notations from \citet{autosubst2}, summarized in Figure~\ref{fig:auxdef}.
Without providing
the full definition of the renaming and substitution functions, it is
impossible to tell the binding structure from the syntax
alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
with the de Bruijn depth of each term, though we note that the
syntax we work with is unscoped and the choice does matter when
we extend our logical relation to open terms in Section~\ref{sec:extension}.

@@

\lang is expressive enough to support large
elimination, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

@@

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction
principle. However, as we will see in Lemma~\ref{lemma:piintroalt}, it
is possible to show the admissibility of \rref{I-PiAlt} a posteriori.

@@

      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (up Suc i G i) { rho } S ]] \text{, then } [[rho i in S]] \\
      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
                       G]]\text{ then there exists some } [[j]] \text{
                       and } [[S]] \text{ such that } [[InterpR j A
                       {rho} S]] \text{ and } [[a {rho} in S]] \\
      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
\end{array}
\]
  \caption{Semantic Typing for \lang}
  \label{fig:semtyping}
\end{figure}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas and leave the remaining cases as exercises for the reader.
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= i : up Suc i G i]]$.
\end{lemma}
\begin{proof}
  TODO: formulate renaming
\end{proof}

@@

Let us first start with some auxiliary lemmas and definitions.
\begin{lemma}[Ext Wn]
  \label{lemma:extwn}
  If $[[wn a i]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn a
  i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

>>>>

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Yiyun Liu}
\orcid{0009-0006-8717-2498}
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \country{USA}
}
\email{liuyiyun@seas.upenn.edu}

\author{Stephanie Weirich}
\orcid{0000-0002-6756-9168}
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \country{USA}
}
\email{sweirich@seas.upenn.edu}

@@

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

%         \mathit{Renaming} \\
%         [[var i {xi}]] & := & [[xi i]] \\
%         [[(Set i) {xi}]] & := & [[Set i]] \\
%         [[Void {xi}]] & := & [[Void]]\\
%         [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
%         [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
%         [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
%         [[Bool {xi}]] & := & [[Bool]] \\
%         [[true {xi}]] & := & [[true]] \\
%         [[false {xi}]] & := & [[false]] \\
%         [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
%         [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
%         [[refl {xi}]] & := & [[refl]] \\
%         [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper. We find de Bruijn
representation advantageous for specification since it leaves very
little ambiguity about the variable freeness side
conditions, making our proof more easily reproducible. Furthermore, as we discuss in
Section~\ref{sec:automation}, de Bruijn representation is much more
amenable to automated reasoning in proof assistants.
We adopt from \citet{autosubst} the notations for simultaneous
renaming, substitution, and other auxiliary definitions related to our
syntax, summarized in Figure~\ref{fig:auxdef}.

\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \mathit{Identity (Ren)} \\
        [[id i]] & := & [[i]] \\ \\
        \mathit{Identity (Subst)} \\
        [[idtm i]] & := & [[var i]] \\ \\
        \mathit{Cons (Ren)} \\
        ([[xi .: n]])([[0]]) & := & [[n]] \\
        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{Cons (Subst)} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{Curried\ Add} \\
        [[up n i]] & := & [[n]] + [[i]] \\ \\
        \mathit{Renaming} \\
        [[a < xi >]] & := & \ldots \\ \\
        \mathit{Up} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[a { rho }  ]] & := & \ldots \\
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

% Without providing
% the full definition of the renaming and substitution functions, it is
% impossible to tell the binding structure from the syntax
% alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
% with the de Bruijn depth of each term, though we note that the
% syntax we work with is unscoped and the choice does matter when
% we extend our logical relation to open terms in Section~\ref{sec:extension}.

@@

\lang is expressive enough to support large
elimination, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

@@

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[F]]$ each time we apply it. Therefore, in
Lemma~\ref{lemma:piintroalt}, we show the admissibility of
\rref{I-PiAlt} a posteriori and use that as a more convenient
introduction rule for the rest of our proofs.

@@

      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
                       G]]\text{ then there exists some } [[j]] \text{
                       and } [[S]] \text{ such that } [[InterpR j A
                       {rho} S]] \text{ and } [[a {rho} in S]] \\
      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
\end{array}
\]
  \caption{Semantic Typing for \lang}
  \label{fig:semtyping}
\end{figure}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas and leave the remaining cases as exercises for the reader.
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  TODO: formulate renaming
\end{proof}

@@

Let us first start with some auxiliary lemmas and definitions.
\begin{lemma}[Ext Wn]
  \label{lemma:extwn}
  If $[[wn a var i]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn a
  var i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}
